Parallelizing AdaBoost and Decision Trees

### 2015-08-21 paper.tex
paper changes

\section{Introduction} Binary and multiclass classification is a fundamental
problem in machine learning and is commonly one of the first concepts to be
studied in an introductory course. In the context of binary classification, a
labeled observation is a coordinate pair $(x, y)$ where $x\in\mathbb{R}^d$ is a
vector of data associated with the observation (feature vector) and $y\in\{1,
-1\}$ is the corresponding label of the observation. For example, $x_i$ might
be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th
individual in a data set while the value of $y_i$ may indicate the individual's
gender. The goal in classification is to accurately predict the label
of a given observation using its feature vector. In this setting, the scenario
of supervised learning involves using a training set of labeled observations
$(x_i, y_i)$ to construct a prediction function $h$, one that takes a feature
vector $\hat{x}$ and gives a prediction on the corresponding label
$h(\hat{x})=\hat{y}$. To check the accuracy of the predictor it is common to use a separate set of test data, i.e. data that hasn't been used for training. There are many methods used to obtain a predictor for classification in supervised
learning, and the class of ensemble methods concerns combining several
predictors together.

@@

The goal of this project is to implement a parallelized version of a commonly
used weak learner, decision trees, to improve the running time of AdaBoost. We
give a short overview of AdaBoost and decision trees in the next section,
describe the implementation of the decision trees, the data sets used and the
results comparing the serial and parallelized execution of the boosting algorithm.

\section{AdaBoost} The beauty of AdaBoost is in using a relatively inaccurate
weak learning algorithm to produce a much more accurate prediction function. In
each round of AdaBoost for $t$ from 1 to $T$, the weak learner selects a
base classifier $h_t: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions
$H$, in particular choosing the one with the least error
$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4) where $D_t(i)$ is
the weight associated with the $i$th sample. The algorithm then re-weights the
training samples in Line 8 to place more emphasis on those that that the chosen
classifier $h_t$ incorrectly predicted, so that in future rounds the base
classifiers are tailored to correcting these mistakes. The predictor that is
returned after all rounds is a linear combination of all $h_t$'s. The
coefficient $\alpha_t$ gives more weight in the final predictor to those rounds
where the classifier performed better. $Z_t$ is a normalization constant. We
give the algorithm below.\\ \\
{\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
\begin{enumerate}
\itemsep1pt \parskip0pt \parsep0pt
\item {\bf for $i=1$ to $n$}
\item \quad $D_1(i) = \frac{1}{n}$
\item {\bf for $t=1$ to $T$}
\item \quad $h_t$ = {\sc WeakLearner}$(S\sim D_t)$ with weighted error $\epsilon_t$
\item \quad $\alpha_t = \frac{1}{2}\log(\frac{1-\epsilon_t}{\epsilon_t})$
\item \quad $Z_t = 2[\epsilon_t(1-\epsilon_t)]^{1/2}$ (normalization factor)
\item \quad {\bf for $i = 1$ to $n$}
\item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
\item $g = \sum_{t=1}{T} \alpha_t h_t$
\item {\bf return} $\mbox{sign}(g)$
\end{enumerate}

@@

\section{Decision Trees}
Decision trees are a well-known method used in classification which partitions
the feature space into rectangular regions and gives each region the most
frequent label among the training set points it contains. In the most common
decision tree algorithms each partition of the feature space is made along
coordinate axes and done in a heirarchical way.  Starting with the entire
space, a line parallel to an axis divides it into two regions and then the same
is done for these resulting regions. This process is repeated many times until
the space is partitioned into several rectangles, like the example in Figure 1
(\cite{HTF}). Each split is made to separate points of the two distinct classes
into different regions as best as possible. Thus the two resulting rectangles
will have a greater {\it purity} in comparison to the parent node.

@@

Since the tree-building process is the most time-consuming portion of AdaBoost,
our goal is to parallelize this. It seems natural to divide amongst processors
the work of finding the dimension in the feature space that best splits the
region and increases the purity for each node in the tree. The datasets we are
handling have several dimensions, on the order of tens to hundreds, so each
processor could be responsible for a group of dimensions in the feature space.
After each processor determines the best dimension to split on amongst it's
group, we can perform a reduction and determine the best split over all
dimensions.

>>>>

\section{Introduction}

\indent Binary and multiclass classification is a fundamental
problem in machine learning and is commonly one of the first concepts to be
studied in an introductory course. In the context of binary classification, a
labeled observation is a coordinate pair $(x, y)$ where $x\in\mathbb{R}^d$ is a
vector of data associated with the observation (feature vector) and $y\in\{1,
-1\}$ is the corresponding label of the observation. For example, $x_i$ might
be a point in $\mathbb{R}^2$ representing the height and weight of the $i$th
individual in a data set while the value of $y_i$ may indicate the individual's
gender. The goal in classification is to accurately predict the label
of a given observation using its feature vector. In this setting, the scenario
of supervised learning involves using a training set of labeled observations
$(x_i, y_i)$ to construct a prediction function $h$, one that takes a feature
vector $\hat{x}$ and gives a prediction on the corresponding label
$h(\hat{x})=\hat{y}$. To check the accuracy of the predictor it is common to use a separate set of test data, i.e. data that hasn't been used for training. There are many methods used to obtain a predictor for classification in supervised
learning, and the class of ensemble methods concerns combining several
predictors together.

@@

The goal of this project is to implement a parallelized version of decision trees, a commonly
used weak learner, in order to improve the running time of AdaBoost. We
give a short overview of AdaBoost and decision trees in the next section,
describe the implementation of the decision trees, the data sets used and the
results comparing the serial and parallelized execution of the boosting algorithm.

\section{AdaBoost}

\indent The beauty of AdaBoost is in using a relatively inaccurate
weak learning algorithm to produce a much more accurate prediction function. In
each round of AdaBoost for $t$ from 1 to $T$, the weak learner selects a
base classifier $h_t: \mathbb{R}^n\rightarrow\{1,-1\}$ from a set of functions
$H$, in particular choosing the one with the least error
$\epsilon_t=\sum_{i=1}^{n}D_t(i)I(h_t(x_i)\neq y_i)$ (Line 4), where $D_t(i)$ is
the weight associated with the $i$th sample. The coefficients $\alpha_t$ and $Z_t$ are calculated, where $\alpha_t$ is used in the ensemble prediction function and $Z_t$ is a normalization constant. The algorithm then updates the
weights in Line 8 to place more emphasis on the samples that $h_t$ misclassified,
so that in future rounds the base classifiers are chosen to correct these mistakes. The predictor that is
returned in Line 9 after $T$ rounds is a linear combination of all $h_t$'s. Note the
coefficient $\alpha_t$ gives more weight in the final predictor to the classifiers from rounds that fared better.
We give the algorithm below.\\ \\
{\sc AdaBoost}: $S = \{(x_i,y_i):i = 1\ldots n\}$
\begin{enumerate}
\itemsep1pt \parskip0pt \parsep0pt
\item {\bf for $i=1$ to $n$}
\item \quad $D_1(i) = \frac{1}{n}$
\item {\bf for $t=1$ to $T$}
\item \quad $h_t$ = {\sc WeakLearner}$(S\sim D_t)$ with weighted error $\epsilon_t$
\item \quad $\alpha_t = \frac{1}{2}\log(\frac{1-\epsilon_t}{\epsilon_t})$
\item \quad $Z_t = 2[\epsilon_t(1-\epsilon_t)]^{1/2}$ (normalization factor)
\item \quad {\bf for $i = 1$ to $n$}
\item \quad \quad $D_{t+1}(i) = \frac{D_t(i)\exp(-\alpha_t y_i h_t(x_i))}{Z_t}$
\item $g = \sum_{t=1}^T \alpha_t h_t$
\item {\bf return} $\mbox{sign}(g)$
\end{enumerate}

@@

\section{Decision Trees}

\indent Decision trees are a well-known method used in classification which partitions
the feature space into rectangular regions and gives each region the most
frequent label among the training set points it contains. In the most common
decision tree algorithms each partition of the feature space is made along
coordinate axes and done in a heirarchical way.  Starting with the entire
space, a line parallel to an axis divides it into two regions and then the same
is done for these resulting regions. This process is repeated many times until
the space is partitioned into several rectangles, like the example in Figure 1
(\cite{HTF}). Each split is made to separate points of the two distinct classes
into different regions as best as possible. Thus the two resulting rectangles
will have a greater {\it purity} in comparison to the parent node.

@@

Since the tree-building process is the most time-consuming portion of AdaBoost,
our goal is to parallelize this. It seems natural to divide amongst processors
the work of finding the dimension in the feature space that best splits the
region and increases the purity for each node in the tree. The datasets we are
handling have several dimensions, on the order of tens to hundreds, so each
processor could be responsible for a group of dimensions in the feature space.
After each processor determines the best dimension to split amongst its
group, we can perform a reduction and determine the best split over all
dimensions.
